---
title: "RmecabKo"
output:
  github_document:
    toc: false
    html_preview: false
---

<!-- README.md is generated from readme.rmd. Edit this file, then render it. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  out.width = "100%"
)
```

[![R-CMD-check](https://github.com/junhewk/RmecabKo/actions/workflows/R-CMD-check.yaml/badge.svg?branch=master)](https://github.com/junhewk/RmecabKo/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/RmecabKo)](https://CRAN.R-project.org/package=RmecabKo)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/RmecabKo)](https://cran.r-project.org/package=RmecabKo)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/RmecabKo)](https://cran.r-project.org/package=RmecabKo)
[![R version](https://img.shields.io/badge/R-%E2%89%A5%203.6.0-276DC3.svg)](https://www.r-project.org/)
[![License: GPL (>= 2)](https://img.shields.io/badge/license-GPL%20(%3E%3D%202)-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)

`RmecabKo` is a Korean text-analysis toolkit for R, built on
[`RcppMeCab`](https://cran.r-project.org/package=RcppMeCab) and `mecab-ko-dic`.
It provides `tidytext`-ready tokenizers, curated Korean stopword data, access
to the KNU sentiment lexicon, friendly user-dictionary management, predicate
lemmatization, keyword extraction, keyword-in-context concordances, and text
normalization.

**RmecabKo is the only actively maintained CRAN package dedicated to Korean
text analysis.** The dedicated alternatives have left CRAN — `KoNLP` was
archived in 2020 and `elbird` (a Kiwi binding) in 2023 — so `RcppMeCab` (the
native engine, by the same author) and `RmecabKo` (this analysis layer on top
of it) are the maintained, `install.packages()`-able path for Korean NLP in R.

## What's new in 0.3.0

- **Tidytext-ready tokenizers.** `token_morph()`, `token_words()`,
  `token_nouns()`, and `token_ngrams()` follow the `tokenizers` contract, so
  they drop straight into `tidytext::unnest_tokens(token = ...)`. New
  `simplify` and `drop_pos` arguments.
- **Korean data.** `stopwords_ko`, a curated POS-tagged stopword table with
  `stopwords_ko_words()` / `stopwords_ko_tags()`, and `lexicon_knu()` for the
  KNU sentiment lexicon.
- **User dictionaries.** `dict_add_words()` and friends register new words from
  R, filling the `mecab-ko-dic` context IDs and final-consonant flag for you.
- **Analysis helpers.** `token_lemma()` (predicate dictionary forms),
  `keywords_tfidf()` / `keywords_textrank()`, `kwic()`, and `text_normalize()`.
- A `vignette("korean-text-analysis")` walks through the full tidy workflow.

For Korean documentation, see
[README_ko.md](https://github.com/junhewk/RmecabKo/blob/master/README_ko.md).

## Why two packages?

The responsibilities are intentionally separated:

| Package | Responsibility |
|---|---|
| `RcppMeCab` | Native MeCab build, R/C++ bindings, parallel analysis, dictionary metadata, and user-dictionary compilation |
| `RmecabKo` | Korean dictionary validation, `tidytext`-ready tokenizers, stopword and sentiment data, user-dictionary management, lemmatization, keyword extraction, KWIC, and text normalization |

This keeps platform-specific native code in one place while allowing
`RmecabKo` to expose Korean-friendly behavior. If the active MeCab dictionary
does not produce Korean POS tags, `RmecabKo` stops with a diagnostic instead of
silently returning Japanese or otherwise incompatible results.

## Installation

Install the stable release from CRAN:

```{r install-cran}
install.packages("RmecabKo")
```

Install the development version from GitHub:

```{r install-github}
install.packages("remotes")
remotes::install_github("junhewk/RmecabKo")
```

`RcppMeCab` is installed as a dependency and supplies the native MeCab
interface. Most Korean-profile installations also include the matching engine
and dictionary. If no Korean dictionary is active, install and select one with:

```{r install-korean-dictionary, eval=FALSE}
RcppMeCab::download_dic("ko")
RcppMeCab::set_dic("ko")
```

A compatible `mecab-ko` engine is required; selecting `mecab-ko-dic` does not
convert a standard Japanese MeCab engine into the Korean engine. Alternatively,
pass the directory of an existing `mecab-ko-dic` installation through
`sys_dic`.

## Quick start

```{r quick-start}
library(RmecabKo)

text <- c(
  first = "한국어 형태소 분석을 합니다.",
  second = "R에서도 빠르게 처리할 수 있습니다."
)

pos(text)
nouns(text)
words(text)
token_morph(text, strip_punct = TRUE)
```

Analysis functions accept a character vector or a list containing one
character value per document. List output always contains one character vector
per input document, following the `tokenizers` contract: supplied document
names are preserved, and unnamed input yields an unnamed list.

Missing documents remain `NA_character_`, while valid empty documents return
`character(0)` from tokenizers and n-gram functions.

## POS tagging

`pos()` is the Korean-validated compatibility layer over `RcppMeCab::pos()` and
`RcppMeCab::posParallel()`.

```{r pos}
pos("한국어 형태소 분석")
pos("한국어 형태소 분석", join = FALSE)
pos(text, format = "data.frame")
pos(text, parallel = TRUE)
```

- `join = TRUE` returns strings such as `"한국어/NNG"`.
- `join = FALSE` returns morphemes with POS tags stored as vector names.
- `format = "data.frame"` exposes the structured output from `RcppMeCab`.
- `parallel = TRUE` uses parallel parsing when multiple documents are supplied.

## Korean tokenizers

The package provides three POS-aware token presets:

| Function | Tokens retained |
|---|---|
| `token_morph()` | All morphemes, optionally restricted with `keep_pos` |
| `token_nouns()` / `nouns()` | Korean POS categories beginning with `N` |
| `token_words()` / `words()` | Categories beginning with `N`, `V`, `M`, or `I`, plus foreign-language tokens tagged `SL` |

```{r tokenizers}
token_morph(text)
token_morph(text, strip_punct = TRUE, strip_numeric = TRUE)
token_morph(text, keep_pos = c("NNG", "NNP"))
token_nouns(text)
token_words(text)
```

Punctuation and numeric filtering is applied to POS-tagged tokens, not to the
raw input string. This avoids accidentally joining text before morphological
analysis. Compound tags such as `VCP+EF` match `keep_pos` when any component is
selected.

## N-grams and skip-grams

`token_ngrams()` creates n-grams after Korean morphological analysis.

```{r ngrams}
token_ngrams(text, n = 2)
token_ngrams(text, n = 1:3, div = "words")
token_ngrams(text, n = 2, skip = 0:1)
token_ngrams(text, n = 2, keep_pos = c("NNG", "NNP"))
token_ngrams(text, n = 2, stopwords = c("분석"), ngram_delim = "_")
```

Important semantics:

- `n` accepts one or more positive integer sizes.
- `skip = 0` creates contiguous n-grams.
- Larger `skip` values are exact gaps: with tokens `a b c d`, a bigram with
  `skip = 1` produces `a c` and `b d`.
- Multiple `n` and `skip` values are returned in requested `n`, `skip`, and
  document-position order.
- Stopwords are boundaries. Terms on opposite sides of a stopword never become
  falsely adjacent.
- Documents shorter than the requested span return `character(0)` safely.

## Custom dictionaries

Teach the analyzer new words from R without hand-writing `mecab-ko-dic` CSV
rows. `RmecabKo` fills in the context IDs and final-consonant flag, compiles the
dictionary through `RcppMeCab`, and activates it for the session:

```{r user-dictionary, eval=FALSE}
dict_add_words(c("은전한닢", "카비봇"), tag = "NNP")
dict_use()
pos("카비봇 출시 소식")
dict_words()
dict_remove_words("카비봇")
```

Pass a data frame for finer control over `reading`, `meaning`, and `cost`, or
supply `left_id`/`right_id` to override the automatic lookup. To inspect the
dictionary currently loaded by MeCab, use `RcppMeCab::dictionary_info()`.

## Tidytext, data, and analysis helpers

The tokenizers follow the `tokenizers` contract, so they drop straight into a
tidy pipeline, and the package ships curated Korean data and analysis helpers:

```{r helpers, eval=FALSE}
library(dplyr)
library(tidytext)

tibble(doc = names(demo_ko), text = demo_ko) |>
  unnest_tokens(word, text, token = token_nouns) |>
  anti_join(data.frame(word = stopwords_ko_words()), by = "word") |>
  count(doc, word, sort = TRUE)

keywords_tfidf(demo_ko, div = "nouns")       # TF-IDF keywords
token_lemma("아침을 먹었다")                  # predicate dictionary forms
kwic(demo_ko, "분석")                         # keyword in context
text_normalize("분석 ㅋㅋㅋㅋ 재밌어요!!!!")  # NFC, width, run squashing
lexicon_knu()                                 # KNU sentiment lexicon
```

See `vignette("korean-text-analysis")` for a full walkthrough.

## Migrating to 0.2.0

Version 0.2.0 removes the duplicated native POS implementation and relies on
`RcppMeCab` for all MeCab integration. Existing public analysis functions are
retained, with these intentional changes:

- `install_mecab()` is deprecated because `RcppMeCab` installs the engine and
  dictionary.
- The old unexported dictionary installer is removed; use
  `RcppMeCab::dict_index()` for user dictionaries.
- Noun extraction once again retains all Korean `N*` categories.
- Token output is consistently list-based and never simplified by `sapply()`.
- Stopwords split n-gram sequences instead of creating artificial adjacency.
- Invalid n-gram sizes and skips now fail clearly instead of risking native
  indexing errors.

## Development

The test suite includes deterministic unit tests, native n-gram tests, and an
end-to-end Korean dictionary integration test. GitHub Actions checks R release
on Linux, macOS, and Windows, R-devel on Linux, and verifies that a Japanese
dictionary is rejected clearly.

```{r development}
devtools::test()
devtools::check(args = c("--as-cran", "--no-manual"))
```

Please report bugs and feature requests in the
[GitHub issue tracker](https://github.com/junhewk/RmecabKo/issues).

## License

`RmecabKo` is licensed under GPL (>= 2). See the package `DESCRIPTION` for
details.
