---
title: "RmecabKo"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

<!-- README.md is generated from readme.rmd. Edit this file, then render it. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  out.width = "100%"
)
```

[![R-CMD-check](https://github.com/junhewk/RmecabKo/actions/workflows/R-CMD-check.yaml/badge.svg?branch=master)](https://github.com/junhewk/RmecabKo/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/RmecabKo)](https://CRAN.R-project.org/package=RmecabKo)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/RmecabKo)](https://cran.r-project.org/package=RmecabKo)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/RmecabKo)](https://cran.r-project.org/package=RmecabKo)
[![R version](https://img.shields.io/badge/R-%E2%89%A5%203.6.0-276DC3.svg)](https://www.r-project.org/)
[![License: GPL (>= 2)](https://img.shields.io/badge/license-GPL%20(%3E%3D%202)-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.html)

`RmecabKo` provides Korean-specific text analysis in R using
[`RcppMeCab`](https://cran.r-project.org/package=RcppMeCab) and
`mecab-ko-dic`. It focuses on a small, dependable API for Korean POS tagging,
noun and content-word extraction, morpheme tokenization, and morpheme-aware
n-grams and skip-grams.

For Korean documentation, see [README_ko.md](README_ko.md).

## Why two packages?

The responsibilities are intentionally separated:

| Package | Responsibility |
|---|---|
| `RcppMeCab` | Native MeCab build, R/C++ bindings, parallel analysis, dictionary metadata, and user-dictionary compilation |
| `RmecabKo` | Korean dictionary validation, Korean POS presets, token filtering, and morpheme n-grams |

This keeps platform-specific native code in one place while allowing
`RmecabKo` to expose Korean-friendly behavior. If the active MeCab dictionary
does not produce Korean POS tags, `RmecabKo` stops with a diagnostic instead of
silently returning Japanese or otherwise incompatible results.

## Installation

Install the stable release from CRAN:

```{r install-cran}
install.packages("RmecabKo")
```

Install the development version from GitHub:

```{r install-github}
install.packages("remotes")
remotes::install_github("junhewk/RmecabKo")
```

`RcppMeCab` is installed as a dependency and supplies the native Korean MeCab
engine and dictionary. A separate system-wide MeCab installation is normally
not required.

## Quick start

```{r quick-start}
library(RmecabKo)

text <- c(
  first = "한국어 형태소 분석을 합니다.",
  second = "R에서도 빠르게 처리할 수 있습니다."
)

pos(text)
nouns(text)
words(text)
token_morph(text, strip_punct = TRUE)
```

Analysis functions accept a character vector or a list containing one
character value per document. List output always contains one character vector
per input document. Supplied document names are preserved; otherwise the input
text is used as the document name.

Missing documents remain `NA_character_`, while valid empty documents return
`character(0)` from tokenizers and n-gram functions.

## POS tagging

`pos()` is the Korean-validated compatibility layer over `RcppMeCab::pos()` and
`RcppMeCab::posParallel()`.

```{r pos}
pos("한국어 형태소 분석")
pos("한국어 형태소 분석", join = FALSE)
pos(text, format = "data.frame")
pos(text, parallel = TRUE)
```

- `join = TRUE` returns strings such as `"한국어/NNG"`.
- `join = FALSE` returns morphemes with POS tags stored as vector names.
- `format = "data.frame"` exposes the structured output from `RcppMeCab`.
- `parallel = TRUE` uses parallel parsing when multiple documents are supplied.

## Korean tokenizers

The package provides three POS-aware token presets:

| Function | Tokens retained |
|---|---|
| `token_morph()` | All morphemes, optionally restricted with `keep_pos` |
| `token_nouns()` / `nouns()` | Korean POS categories beginning with `N` |
| `token_words()` / `words()` | Categories beginning with `N`, `V`, `M`, or `I`, plus foreign-language tokens tagged `SL` |

```{r tokenizers}
token_morph(text)
token_morph(text, strip_punct = TRUE, strip_numeric = TRUE)
token_morph(text, keep_pos = c("NNG", "NNP"))
token_nouns(text)
token_words(text)
```

Punctuation and numeric filtering is applied to POS-tagged tokens, not to the
raw input string. This avoids accidentally joining text before morphological
analysis. Compound tags such as `VCP+EF` match `keep_pos` when any component is
selected.

## N-grams and skip-grams

`token_ngrams()` creates n-grams after Korean morphological analysis.

```{r ngrams}
token_ngrams(text, n = 2)
token_ngrams(text, n = 1:3, div = "words")
token_ngrams(text, n = 2, skip = 0:1)
token_ngrams(text, n = 2, keep_pos = c("NNG", "NNP"))
token_ngrams(text, n = 2, stopwords = c("분석"), ngram_delim = "_")
```

Important semantics:

- `n` accepts one or more positive integer sizes.
- `skip = 0` creates contiguous n-grams.
- Larger `skip` values are exact gaps: with tokens `a b c d`, a bigram with
  `skip = 1` produces `a c` and `b d`.
- Multiple `n` and `skip` values are returned in requested `n`, `skip`, and
  document-position order.
- Stopwords are boundaries. Terms on opposite sides of a stopword never become
  falsely adjacent.
- Documents shorter than the requested span return `character(0)` safely.

## Custom dictionaries

Compiling a user dictionary remains the responsibility of `RcppMeCab`:

```{r user-dictionary}
RcppMeCab::dict_index(
  dic_csv = "user-words.csv",
  out_dic = "user-words.dic",
  dic_dir = getOption("mecabSysDic")
)

user_dictionary <- normalizePath("user-words.dic")
pos("새로 등록한 단어", user_dic = user_dictionary)
```

A Korean proper-noun entry follows the `mecab-ko-dic` CSV layout, for example:

```text
새단어,,,,NNP,*,F,새단어,*,*,*,*
```

Use full paths for `sys_dic` and `user_dic`. To inspect the dictionary currently
loaded by MeCab, use:

```{r dictionary-info}
RcppMeCab::dictionary_info()
```

## Migrating to 0.2.0

Version 0.2.0 removes the duplicated native POS implementation and relies on
`RcppMeCab` for all MeCab integration. Existing public analysis functions are
retained, with these intentional changes:

- `install_mecab()` is deprecated because `RcppMeCab` installs the engine and
  dictionary.
- The old unexported dictionary installer is removed; use
  `RcppMeCab::dict_index()` for user dictionaries.
- Noun extraction once again retains all Korean `N*` categories.
- Token output is consistently list-based and never simplified by `sapply()`.
- Stopwords split n-gram sequences instead of creating artificial adjacency.
- Invalid n-gram sizes and skips now fail clearly instead of risking native
  indexing errors.

## Development

The test suite includes deterministic unit tests, native n-gram tests, and an
end-to-end Korean dictionary integration test. GitHub Actions checks R release
on Linux, macOS, and Windows, R-devel on Linux, and verifies that a Japanese
dictionary is rejected clearly.

```{r development}
devtools::test()
devtools::check(args = c("--as-cran", "--no-manual"))
```

Please report bugs and feature requests in the
[GitHub issue tracker](https://github.com/junhewk/RmecabKo/issues).

## License

`RmecabKo` is licensed under GPL (>= 2). See the package `DESCRIPTION` for
details.
